Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors

### 2014-10-30 paper_v1.Rnw
wrote S1 discussion, edited abstract

\title{Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses}

@@

\begin{abstract}
Research on attrition has a long tradition in the field of education. Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation. A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses suggested time to be a critical obstacle and [other important take-aways from Study 1] (Study 1; $N = XX,XXX$). A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset. Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%). A low capacity for volitional control was identified as a large contributor to this issue. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

@@

To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following five factors emerged: first, general time issues (course requires too much time; can't keep up with deadlines); second, late start (starting course late); third, difficulty (too advanced, exam too difficult); fourth, format and content (didn't enjoy online format, confusing videos, materials not explained well), and fifth, expectations (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable). Only the technical difficulties item failed to load onto any factor.

@@

<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)

@@

Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward. Expectancy-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain. In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks. Because learners can join or leave MOOCs freely. Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts. Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification. So what does lack of time for the course mean? In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".

@@

>>>>

\title{Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors}

@@

\begin{abstract}
Research on attrition has a long tradition in the field of education. Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation. A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1). Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence. A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset. Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%). A low capacity for volitional control was identified as a large contributor to this issue. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

@@

To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em  difficulty} (too advanced, exam too difficult); fourth, {\em format and content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable). Only the technical difficulties item failed to load onto any factor.

@@

<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)

@@

A number of striking findings emerged from the preceeding analyses. We found strong evidence for gender inequalities in attrition and, to some extent, performance. Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship. The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women. Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched. We also identified a large achievement gap in terms of grades between learners from these continents. The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings. It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.

Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end. Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations. General time issues were overall the most pronounced.

Learners may disengage from the course when they become too occupied by higher-priority tasks. Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain. A low capacity for volitional control \cite{corno2001volitional} could lead learners misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.

@@

We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control. In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
